Learning a functional distance between organizations with unsupervised embeddings of scientific mobility

### 2020-07-23 paper/current/main.tex

Prestige hierarchies underpin the dynamics of scientific mobility~\autocite{deville2014career, clauset2015hierarchy}, but are these hierarchies captured in the embedding space, to which no explicit information on prestige is given?
We test this by exploiting the geometric properties of the embedding space with SemAxis~\autocite{an2018semaxis}, an approach originally designed to characterize the semantics of words in word embeddings.
Here, we use SemAxis to operationalize the abstract notion of academic prestige, defining each pole of the axis using known high- and low-ranked universities. 
We define prestige using the Times Ranking of World Universities (and using research impact from the Leiden Rankings~\autocite{waltman2012leidenrankings} in Supplementary Materials), including all ranked universities in the United States;
the high-ranked pole is defined as the mean vector of the top five U.S. universities according to the rankings, whereas the low-ranked pole is defined using the five bottom-ranked and geographically matched (by U.S. census region) universities.
A geographic axis is also added to aid the visualization and reinforce the importance of geography, defined as the mean vector of all universities in California and then Massachusetts.

Prestigious universities such as Colombia, Stanford, MIT, Harvard, and Rockefeller are positioned towards the top of the axis (Fig.~\ref{fig:semaxis}a).
Universities at the bottom of this axis tend to be regional universities with lower national profiles (yet still ranked by Times Higher Education) and with more emphesis on teaching, such as Barry Univeristy and California State University at Long Beach.
SemAxis also offers a new means of reasoning about the prestige of organizations typically excluded from rankings such as government organizations (Fig.~\ref{fig:semaxis}b), research institutes (Fig.~\ref{fig:semaxis}c), and teaching colleges (Fig.~\ref{fig:semaxis}d).

@@

The embedding space captures prestige hierarchies that can be used to reconstruct academic rankings. 
We derive a ranking for universities using the SemAxis prestige axis in Fig.~\ref{fig:semaxis}a and compare them to traditional university rankings.
The SemAxis ranking is defined from the position of organizations along the prestige axis, with the top-ranked being the furthest towards the high-ranked end of the axis.
We find that the SemAxis ranking is correlated with the Times ranking (Spearman's $\rho = 0.73$, Fig.~\ref{fig:semaxis}e).
The strongest outliers that were ranked more highly in the Times Ranking than the SemAxis ranking tend to be large state universities such as Arizona State University and  University of Florida;
Those ranked higher in the SemAxis ranking tend to be relatively-smaller universities such as the University of San Francisco and University of Maryland Baltimore County that are close to other prestigious universities.

The correlation between SemAxis and the Times ranking is robust with the number of organizations used to define the axes (Fig. SI~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities are significantly correlated with the Times Ranking (Spearman's $\rho = 0.46$, Fig.~\ref{fig:supp:semaxis_compare}a).
For reference, the SemAxis-derived ranking achieves correlation with the Times Ranking comparable to that between the Times ranking and the Leiden Ranking~\autocite{waltman2012leidenrankings}, a common and bibliometrically-informed university ranking (Spearman's $\rho = 0.87$, Fig.~\ref{fig:semaxis}f).
This suggests that the organization embedding captures the latent prestige hierarchy inherent to academic mobility, even without providing any explicit information on prestige, with a high degree of resolution and reliability~\autocite{deville2014career, clauset2015hierarchy}.

What is it about the organization embedding that captures prestige so effectively?
We find that the length of organization vectors---that is, the distance from the origin of the vector space---provides provides insight into the characteristics of organizations. 
Up to a point, the length of U.S. organization's vectors tends to increase proportional to its number of researchers (both mobile and non-mobile) who published with that organization;
these organizations are primarily teaching-focused institutions, agencies, and hospitals that either are not ranked or have a low ranking. 
However, at around 1,000 researchers, the length of the vector decreases with increase in the number of researchers;
organizations here are primarily research-intensive and prestigious universities with higher rankings, research outputs, R\&D funding, and doctoral students (Fig.~\ref{fig:supp:length_vs_metaInfo}).
This same convex-curve repeats irrespective of country, such as China and Germany, though the curve may be shifted based on the typical size and vector length of their organizations (Fig.~\ref{fig:length}b).

This pattern has been observed in applications of neural embedding to natural language, wherein the length of word vectors were found to represent word "significance"~\autocite{schakel2015measuring}
Here, interpret the vector length instead as a measure of an organization's connectedness to other organizations given its size, where a connection is defined as a non-zero flow between organizations.
for example, organizations with a small L2 norm tend to also have a lower number of connections given their size (Fig.~\ref{fig:length}b, inset, $R^{2} = 0.17$). 
Large and prestigious universities can be considered as having a shorter vector length than expected given their size, reflecting their ability to attract researchers from many places, to place graduates in positions around the world, and the capacity to encourage mobility by offering temporary and visiting researcher positions. 
At the national level, the relative position of the convex-curves can reveal characteristics of their scientific organizations—for example, compared to the United States, Chinese organizations have a longer typical vector length (Fig.~\ref{fig:length}b) reflecting lower overall connectedness in the embedding space. 
Prestige is a key component academic mobility—by encoding information relating to prestige, neural embeddings offer a means of better representing functional distance between organizations and capturing the global landscape of mobility.

%
% Figure - SemAxis results
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/Math/boomerang.pdf}
	\caption{
		\textbf{Length of organization vectors captures connection diversity of organizations.}
		\textbf{a.} L2 norm of organization vectors compared to their number of researchers for U.S. universities.
		The L2 norm corresponds to the distance from the origin of the vector space.
		Color indicates the rank of the university from the Times ranking, with 1 being the highest ranked university.
		Uncolored points are universities not listed on a Times ranking.
		A convex-shape emerges, wherein larger universities tend to more distant from the origin (large L2 norm), however the more prestigious the university, the closer it is to the origin (small L2 norm). 
		\textbf{b.} This convex pattern repeats globally for all organizations across all countries, shown for the United States, Germany, China, and others,
		The loess regression lines are shown for each selected country, and for the aggregate of remaining countries. 
		The L2 norm is correlated weakly with the number of connection divided by the number of researchers in organization, which can be interpreted diversity of the connection compared to their size ($R^{2} = 0.17$, \textbf{inset}).
		The red line is the line of the best fit.
	}
	\label{fig:length}
\end{figure}

>>>>

Prestige hierarchy is known to underpin the dynamics of scientific mobility~\autocite{deville2014career, clauset2015hierarchy}, but can this hierarchy be encoded in the embedding space, to which no explicit information on prestige is given?
We test this by exploiting the geometric properties of the embedding space with SemAxis~\autocite{an2018semaxis}, an approach originally designed to characterize the semantics of words in word embeddings.
Here, we use SemAxis to operationalize the abstract notion of academic prestige, defining each pole of the axis using high- and low-ranked universities. 
We define prestige using the Times Ranking of World Universities (and using research impact from the Leiden Rankings~\autocite{waltman2012leidenrankings} in Supplementary Materials), including all ranked universities in the United States;
the high-ranked pole is defined as the mean vector of the top five U.S. universities according to the rankings, whereas the low-ranked pole is defined using the five bottom-ranked and geographically matched (by U.S. census region) universities.
A geographic axis is also added to aid the visualization and reinforce the importance of geography, defined as the mean vector of all universities in California and then Massachusetts.

Prestigious universities such as Colombia, Stanford, MIT, Harvard, and Rockefeller are positioned towards the top of the axis (Fig.~\ref{fig:semaxis}a).
Universities at the bottom of this axis tend to be regional universities with lower national profiles (yet still ranked by Times Higher Education) and with more emphasis on teaching, such as Barry University and California State University at Long Beach.
SemAxis also offers a new means of reasoning about the prestige of organizations typically excluded from rankings such as government organizations (Fig.~\ref{fig:semaxis}b), research institutes (Fig.~\ref{fig:semaxis}c), and teaching colleges (Fig.~\ref{fig:semaxis}d).

@@

The embedding space captures the prestige hierarchy and can be used to reconstruct academic rankings. 
We derive a ranking for universities using the SemAxis prestige axis in Fig.~\ref{fig:semaxis}a and compare them to traditional university rankings.
The SemAxis ranking is defined from the position of organizations along the prestige axis, with the top-ranked being the furthest towards the high-ranked end of the axis.
We find that the embedding-based ranking is strongly correlated with the Times ranking (Spearman's $\rho = 0.73$, Fig.~\ref{fig:semaxis}e).
The strongest outliers that were ranked more highly in the Times Ranking than the SemAxis ranking tend to be large state universities such as Arizona State University and  University of Florida;
Those ranked higher in the SemAxis ranking tend to be relatively-smaller universities such as the University of San Francisco and University of Maryland Baltimore County that are close to other prestigious universities.

The correlation between SemAxis and the Times ranking is robust with the number of organizations used to define the axes (Fig. SI~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities are significantly correlated with the Times Ranking (Spearman's $\rho = 0.46$, Fig.~\ref{fig:supp:semaxis_compare}a).
For reference, the SemAxis-derived ranking achieves correlation with the Times Ranking comparable to that between the Times ranking and the Leiden Ranking~\autocite{waltman2012leidenrankings}, a well-known and bibliometrically-informed university ranking (Spearman's $\rho = 0.87$, Fig.~\ref{fig:semaxis}f).
This suggests that the organization embedding captures the latent prestige hierarchy inherent to academic mobility, even without providing any explicit information on prestige, with a high degree of resolution and reliability~\autocite{deville2014career, clauset2015hierarchy}.

What is it about the organization embedding that captures prestige so effectively?
We find that the size (L2 norm) of the organization embedding vectors provides insights into the characteristics of organizations. 
Up to a point, the size of U.S. organization's vectors tends to increase proportional to its number of researchers (both mobile and non-mobile) who published with that organization;
these organizations are primarily teaching-focused institutions, agencies, and hospitals that either are not ranked or have a low ranking. 
However, at around 1,000 researchers, the size of the vector decreases as the number of researchers increases;
organizations here are primarily research-intensive and prestigious universities with higher rankings, research outputs, R\&D funding, and doctoral students (Fig.~\ref{fig:supp:length_vs_metaInfo}).
A similar pattern has been observed in applications of neural embedding to natural language, wherein the length of word vectors were found to represent word "significance"~\autocite{schakel2015measuring}.
If a word appears frequently in a specific context, the norm of the word vector is likely to be large, because a large norm is associated with how predictably the word occurs given a context.
However, if the word in question is universal, appearing frequently in many different contexts, it cannot have a large norm as it lacks a strong association with a particular context. 
Likewise, an organization such as Harvard with a small norm appears in many contexts, that is, alongside many different organizations in affiliation trajectories. 
The concavity emerges in part from the relationship between the size of the vector and the expected connectedness of the organization given its size (Fig.~\ref{fig:length}a, inset, $R^{2} = 0.17$).
Large, prestigious, and well-funded research universities such as Princeton and Harvard have smaller vector sizes because they appear in many different contexts compared to more teaching-focused organizations such as NY Medical College, and University of Michigan at Flint. 
Some universities, such as the University of Alaska at Fairbanks, have surprisingly small vectors, perhaps as a result of its remote location and unique circumstances.

This same convex-curve repeats irrespective of country, such as China and Germany, though the curve may be shifted based on the typical size and vector length of their organizations (Fig.~\ref{fig:length}b). 
The relative position of these curves can reveal characteristics about the scientific systems in each country. 
For example,

This 
Here, interpret the vector length instead as a measure of an organization's connectedness to other organizations given its size, where a connection is defined as a non-zero flow between organizations.
for example, organizations with a small L2 norm tend to also have a lower number of connections given their size (Fig.~\ref{fig:length}b, inset, $R^{2} = 0.17$). 
Large and prestigious universities can be considered as having a shorter vector length than expected given their size, reflecting their ability to attract researchers from many places, to place graduates in positions around the world, and the capacity to encourage mobility by offering temporary and visiting researcher positions. 
At the national level, the relative position of the convex-curves can reveal characteristics of their scientific organizations—for example, compared to the United States, Chinese organizations have a longer typical vector length (Fig.~\ref{fig:length}b) reflecting lower overall connectedness in the embedding space. 
Prestige is a key component academic mobility—by encoding information relating to prestige, neural embeddings offer a means of better representing functional distance between organizations and capturing the global landscape of mobility.

%
% Figure - SemAxis results
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/Math/boomerang.pdf}
	\caption{
		\textbf{Size of organization embedding vectors captures connection diversity of organizations.}
		\textbf{a.} Size (L2 norm) of organization embedding vectors compared to their number of researchers for U.S. universities.
		Color indicates the rank of the university from the Times ranking, with 1 being the highest ranked university.
		Uncolored points are universities not listed on a Times ranking.
		A convex-shape emerges, wherein larger universities tend to more distant from the origin (large L2 norm), however the more prestigious the university, the closer it is to the origin (small L2 norm). 
		The L2 norm is correlated weakly with the number of connection divided by the number of researchers in organization, which can be interpreted diversity of the connection compared to their size ($R^{2} = 0.17$, \textbf{inset}).
		The red line is the line of the best fit, and 99\% confidence intervals are shown. 
		\textbf{b.} This convex pattern repeats globally for all organizations across all countries, shown for the United States, China, and others,
		The loess regression lines are shown for each selected country, and for the aggregate of remaining countries. 
	}
	\label{fig:length}
\end{figure}
